---
title: "Dimensionality reduction with Uniform Manifold Approximation and Projection"
date: "2023-02-07"
categories: [dimensionality reduction, UMAP, EDA]
description: 'Using UMAP with R'
draft: true
---

Uniform Manifold Approximation and Projection, [UMAP](https://arxiv.org/abs/1802.03426) for short, is algorithm which can be used to reduce the number of dimensions in a data set. Two packages [umap](https://cran.r-project.org/web/packages/umap/vignettes/umap.html) and [uwot](https://cran.r-project.org/web/packages/uwot/index.html), which seems to be used by {tidymodels}.


## Using the built-in data `swiss`

Let's explore the use of UMAP and PCA by using the built-in `swiss` data.

```{r}
# load the built-in data
data(swiss)

head(swiss)
```

 ### Dimensionality reduction via PCA
 
```{r}
library(tidyverse)
theme_set(theme_minimal())

# Perform PCA
pca.out <- prcomp(swiss, scale. = TRUE)

municipalities <- row.names(pca.out$x) %>% 
  as_tibble() %>% 
  rename(municipality = value)
# visualize first two principal components
pca.out$x %>% 
  as_tibble() %>% 
  bind_cols(municipalities) %>% 
  ggplot(aes(x = PC1, y = PC2, label = municipality)) +
    geom_point() +
    geom_text(check_overlap = TRUE)

```
 


Julia Silge has a nice [blog post](https://juliasilge.com/blog/cocktail-recipes-umap/) comparing the use of PCA and UMAP via the tidymodels approach.